---
title: "Estimate seroprevalence using a simple and not-so-simple approach"
output: html_notebook
---

In this notebook, I estimate seroprevalence using two different approaches. The first, simple approach simply combines poststratification and the Rogan Gladen correction to account for test imperfections.  The second approach uses a modified versio of multi-level regression and poststratification to more accurately model the data. I fit the MRP approach using both native Stan and the BRMS package.

For more information the two approaches, see my [blog] (ww.dougjohnson.in). 

Nearly all of the code to generate the simulated data is from the R vignette [MRP with rstanarm](https://cran.r-project.org/web/packages/rstanarm/vignettes/mrp.html) by Lauren Kennedy and Jonah Gabry. (Reproduced with their permission.)  Much of the Stan MRP code borrows from the code in the user guide.

## Generate simulated data

Load the relevant libraries and generate simulated data using a slightly modified version of Kennedy and Gabry's code.

```{r setup, message=FALSE}
library(tidyverse); library(rstan); library(glue); library(brms); library(caret)
path <- "C:/Users/dougj/Documents/code/sero_prevalence2"
source(file.path(path, "functions", "simulate_mrp_data_w_noise_random.R"))

sero_sim <- simulate_mrp_data(n=1000)
sample <- sero_sim[['sample']]
poststrat <- sero_sim[['poststrat']]
true_popn <- sero_sim[['true_popn']]

# Data from evaluation of antibody test on samples with disease and negative controls 
tp <- 90  # true positives - disease samples with correct test output
fn <- 10  # false negatives - disease samples with incorrect test output
tn <- 95  # true negatives - negative controls with correct test output
fp <- 5   # false positives - negative controls with incorrect test output


#set.seed(8675309)
se <- rbeta(1, tp, fn)
sp <- rbeta(1, tn, fp)
glue("Sensivity: {round(se, 2)}")
glue("Specificity: {round(sp, 2)}")

# Create binary vectors for whether the antibody test would give the correct result
# (Creating two vectors saves time because the assignment can then be easily vectorized)
pos_correct <- rbinom(n = 1000, size = 1, prob = se)
neg_correct <- rbinom(n = 1000, size = 1, prob = sp)

sample$test <- sample$cat_pref*pos_correct+(1-sample$cat_pref)*(1-neg_correct)

true_popn_pref <- sum(true_popn$cat_pref * poststrat$N) / sum(poststrat$N)
glue("The true prevalence is {round(true_popn_pref, 3)}")
```

## Estimate seroprevalance using the simple approach

I first estimate the overall seroprevalence using the really simple approach.  Note that I first collapse down to a manageable number of strata.  Otherwise, there are many strata for which we have no observations in the sample dataset.

```{r simple}
# collapse the sample and poststrat data on state and age
sample_collapsed <- sample %>% 
  group_by(male, eth, income) %>% 
  summarize(pt_total = sum(cat_pref), pa_total = sum(test), n = n())

poststrat_collapsed <- poststrat %>% 
  group_by(male, eth, income) %>% 
  summarize(N = sum(N))

# merge sample data with poststrat collapsed
res <- sample_collapsed %>% left_join(poststrat_collapsed, by = c("male", "eth", "income"), keep.all = TRUE)

# calculate pt_simple
NT <- sum(res$N)
pa_simple <- sum(res$pa_total*res$N*(1/res$n))/NT
se_hat <-  tp/(tp+fn)
sp_hat <- tn/(tn+fp)
pt_simple <- (pa_simple+sp_hat-1)/(se_hat+sp_hat-1)
glue("The simple estimate of prevalence is {round(pt_simple, 3)}")
```

## Esimate seroprevalence using modified MRP and Stan

Next, estimate seroprevalence using MRP and native Stan.

```{r mmrp_stan}

# Create a new variable that is 1+male since Stan expects sex to be {1,2}^N
poststrat$sex <- poststrat$male+1
ps_reordered <- poststrat[c('sex', 'age', 'income', 'state', 'eth', 'N')] %>% 
  arrange(eth, state, income, age, sex)

# Create an array using contents of ps_reordered to pass to Stan
P <- array(data = ps_reordered$N, 
             dim=head(as.numeric(lapply(ps_reordered, function(x) length(unique(x)))),-1), 
             dimnames=lapply(ps_reordered[,1:(length(ps_reordered)-1)], unique))

# check that P was assigned correctly by looking at the value of N for a random row
dim(P)
P[1, 3, 2, 45, 2]
ps_reordered %>% filter((sex == 1) & (age == 3) & (income == 2) & (state == 45) & (eth == 2))

# Gather the data in a list
mmrp_data <-  list(N = nrow(sample),
                   ng_age = length(unique(ps_reordered$age)),
                   ng_income = length(unique(ps_reordered$income)),
                   ng_state = length(unique(ps_reordered$state)),
                   ng_eth = length(unique(ps_reordered$eth)),
                   sex = sample$male+1,
                   age = sample$age,
                   income = sample$income,
                   eth = sample$eth,
                   state = sample$state, 
                   y = sample$cat_pref,
                   P = P,
                   tp = tp,  
                   fn = fn,
                   tn = tn,
                   fp = fp)

mmrp_fit <- stan(
  file = file.path(path, "stan_code", "mrp_with_uncertainty.stan"),
  data = mmrp_data, 
  iter = 1000, chains = 3
  )

# Inspect Stan output and compare to the true population figures and the unweighted sample figures.
mmrp_draws <- extract(mmrp_fit)
se_posterior <- mean(mmrp_draws$se)
sp_posterior <- mean(mmrp_draws$sp)
pt_pop_posterior <- mean(mmrp_draws$phi)

glue("The not-so-simple estimate of prevalence is {round(pt_pop_posterior, 3)}")
```

## Estimate seroprevalance using the not-so-simple approach and BRMS

Next, I estimate the same model using BRMS.  (Well, almost the same. The BRMS model does not account for the uncertainty in estimation of sensitivity and specificity.)

```{r mmrp_brms, message=FALSE}
# Define the custom family to add to BRMS
sero_noise <- custom_family(
  "sero_noise", 
  links = c("logit"), 
  type = "int"
)

stan_funs <- glue("
  real sero_noise_lpmf(int y, real mu) {{
    real pt;
    pt = mu*{tp/(tp+fn)}+(1-mu)*({1-(tn/(tn+fp))});
    return bernoulli_lpmf(y | pt);
  }}
  real sero_noise_rng(real mu) {{
    return bernoulli_rng(mu);
  }}
")

stanvars <- stanvar(scode = stan_funs, block = "functions")

fit2 <- brm(
  test | trials(1) ~ factor(male) + factor(male) * factor(age) + 
    (1 | state) + (1 | age) + (1 | eth) + (1 | income),
  family = sero_noise,
  data = sample, stanvars = stanvars,
  save_all_pars = TRUE
)

# Save the stan code and data passed to Stan for further inspection
cat(stancode(fit2), file = file.path(path,"stan_code","brms_code2.stan"))
model_data2 <- standata(fit2)

```

Next, I generate predictions using the fitted BRMS object.

```{r gen_predictions}
posterior_predict_sero_noise <- function(i, draws, ...) {
  draws$dpars$mu[, i]
}

# test out the predict function on existing data
test <- predict(fit2)


# Generate predictions using the poststrat data
posterior_prob <- predict(fit2, newdata = poststrat, allow_new_levels = TRUE)

# The posterior predict function uses the predicted value of the prevalence. This is Ok as long as we are just seeking point estimates. To fully account for uncertainty, we should draw from binom(N, p = posterior_prob[i])
poststrat_prob <- posterior_prob[,1] %*% poststrat$N / sum(poststrat$N)

glue("The not-so-simple estimate of prevalence using BRMS is {round(poststrat_prob, 3)}")

```

